Comment: Inferring Trade Costs from Trade Booms and Trade Busts

### 2016-07-28 Paper/Manuscript.tex
jacks et / jacks and

This is potentially an important result that sheds light on the globalization processes. However, that result is actually driven by an \textit{ad hoc}
method of aggregation that captures structure effects. The inferred difference between the two globalization comes from the fact that the dyads with the 
fastest growing trade in the first wave of globalization start with very small 
trade. This is not the case in the second wave of globalization.
Using a three-countries version of \cite{AW2003}'s model for
the aggregation of bilateral trade costs shows that there was no 
difference in nature between the two globalizations. More generally, we
cast doubt on the possibility to distinguish the effect of trade costs and economic expansion through an approach that relies only on the study of trade
flows. Deducing trade costs from trade flows, and then using
trade costs to explain trade flows is essentially a circular
reasoning. Therefore, Jacks and al.'s approach cannot be an
alternative to traditional investigations of impediments to
trade, such as commodity price gaps.

This paper starts by presenting Jacks and al.'s approach of the
measure of trade costs, and insists on its relevance at the
bilateral level to control for multilateral trade barriers in
gravity regressions (section \ref{1}). It then highlights that the
result on a difference of nature between the two globalizations
is paradoxical since it cannot be deduced from a comparison of the
evolution of openness ratios. Section \ref{2} shows that the
conclusion is only driven by the authors' \textit{ad hoc} aggregation
method. We propose a microfounded way to aggregate trade costs
and the puzzle fades away. Section \ref{3} explores the reasons
why Jacks and al.'s aggregation technique ends up providing
different results for the two globalizations. We argue that
what Jacks and al. misleadingly attribute to unequal trade costs
decreases between the two globalizations is instead a
difference in the distribution of trade growth over trading
partners.

@@

Jacks and al. first depart from \cite{AW2003} by eliminating the
multilateral resistance variable ($P_i P_j$) from the gravity equation.
Following \cite{NOVY}, bilateral trade barriers are not
considered relative to multilateral trade barriers modelized by
the price index, but to domestic trade costs. In this case,
trade flows are no longer compared to outputs, but to internal
trade $x_{ii}$\footnote{Due to data limitations, the authors
use the relation $x_{ii}=GDP_i-EXPORTS_i$ to get internal
trade. We follow them. Concerns about the fact that GDP is
measured in added-value and exportations as gross value are
addressed in appendix B of \cite{JMN2011}.}. The equation above
becomes:

@@

Our point is that a method that uses trade flows
alone can teach us much about their determinants. When
departing from \cite{AW2003}'s multilateral resistances, the
arbitrage condition Jacks and al. base their computation on is an
equality between a Marginal Rate of Substitution (MRS) and a
price ratio for a CES utility function, or more precisely the
product of two such equalities, one from the program of each
trade partner. Appendix \ref{A1} derives equation (\ref{keyeq})
this way without passing through multilateral resistance. The
inference of trade costs from trade flows is tantamount to
using the relation between quantities and prices given by a
demand curve derived from fully specified exogenous
preferences.

@@

Since the trade costs measure is assimilated to trade flows
relative to domestic activity, they simply restate
 an information already contained in any measure of
trade flows relative to economic size, like an openness ratio
defined for instance as the ratio of exportations to GDP.
However the main conclusion in Jacks and al.'s article, namely
that the first globalization was driven more by the decrease in
trade costs than the second one, requires the absence of circularity.
Indeed, when we decompose the level of
exportations of a country as the product of its GDP and its
openness ratio, we find results that are different from theirs (see table \ref{OR}).\footnote{We follow
the authors in interpreting log differences as percentages. However, one should keep in mind that given the
size of the changes, this is a very inexact approximation.} The increase in exportations is mainly explained by GDP growth for the large majority of the 27 countries in the
data. On average, this decomposition attributes 74\% (183/246) of
the growth in trade to the increase in GDP in the second
globalization and 60\% in the first globalization. This is very much in contrast to 
Jacks and al.\footnote{Contributions of growth in income
similarity and of change in multilateral factors are negative.} :

@@

Jacks and al.'s conclusion on a difference of nature between the
two globalizations is based on an aggregate trade costs
measure that provides a summary statistic of the
evolution of trade costs across all dyads in the sample, as
well as for different regions of the world. To move from
bilateral costs $\tau_{ij}$ to an aggregate measure of trade
costs, the authors use an arithmetic mean over dyads, weighted
by the sum of the GDP of the two trade partners.

@@

Let us note $\upsilon_{i}$ the corresponding trade cost faced
by country $i$ with all its trading partners in the data. Its
expression is given by Jacks and al.'s key equation
(\ref{keyeq}) applied to a three-countries model:

@@

Using this measure of trade costs, we provide the same
decomposition in four terms of the increase in trade flows
during the two globalizations as in the authors' article for
the countries with the most trade partners available in the
data (France (24), the UK (25), the USA (23)). We also provide
the unweighted and end-of-period-GDP-weighted averages for all
countries in the sample. Results are displayed in tables
\ref{glob1} and \ref{glob2} along with the results with the
Jacks and al's method.\footnote{They include unweighted and
end-of-period-GPD-weighted averages over country dyads of
bilateral trade costs, such as provided in \cite{JMN2011}. We
also display for France, the UK and the USA the trade cost
measure averaged over trading partners faced by one country, as
the results are presented (with a smaller data set) in
\cite{JMN2008, JMN2010}. We also provide both averages of this
measure over all countries in the data set.}\\

@@

Tables \ref{glob1} and \ref{glob2} highlight how much the decomposition
between the decrease in trade costs and the income growth depends on the aggregation technique. When we use the microfounded aggregation method,\footnote{We selected the (GDP-weighted) average of our measure in order
to allow a clear comparison with Jacks and al.'s results. There
is of course no rational for such a summary statistic, but the
results for France, the UK and the USA assure that the main
conclusion of this exercice does not depend on averaging over
countries.} 
growth in output is the main driving force behind both waves of growth in
international trade, contributing to about 56\% (203/365) in
the first wave and 66\% (361/545) in the second one. It is no coincidence
if these results are very similar to the ones in table \ref{OR}: 60\% (71/107) in the first globalization and 74\% (183/246) in the second one.
Insofar as a measure of
trade costs is defined residually as everything that explains
differences between domestic and international trade flows, it
is bound to reword the information contained in an openness
ratio.

@@

Our explanation of why Jacks and al.'s aggregate measure of
trade costs yields different conclusion 
from table \ref{OR} is because it is sensitive to structure
effects. To explain this idea, and for clarity purposes, let us consider
a word where partners are symmetric (domestic trade is equal in $i$ and $j$ and $x_ij$=$x_ji$). From equation \ref{keyeq}, we have:

@@

Trade costs imply that prices are specific to the country of
consumption. We note $p_k$ the supply price of the producer in
country $k$ net of trade costs, and $p_{ki}$ the price of
region $k$ goods for region $i$ consumers. We define
$t_{ki}=p_{ki}/p_{k}$ the trade costs factor between $k$ and
$i$. Jacks and al.'s variable of trade costs between $k$ and
$i$, $\tau_{ki}$ is then defined as the trade costs factor
between $k$ and $i$ relative to the domestic trade costs factor
$t_{ii}$: $\tau_{ki}=t_{ki}/t_{ii}-1$. In all that follows, we
use the same notations $\tau_{ki}$ for the geometric average of
$\tau_{ki}$ and $\tau_{ik}$. Symmetry of bilateral trade costs
does not need to be assumed.

@@

This is a static result. We are interested in the dynamic
behavior of trade flows. But Jacks and al.'s measure also
overweights initially small trade partners in terms of the
increase in trade flows. This can be seen by calculating the
elasticities to the importations from country $k$ ($a_k$) of
the arithmetic mean ($m$) and of Jacks and al.'s mean
($m_{\sigma}$) (keeping importations from other trade partners
$a_j$ constant). The elasticity of a generic mean $m_\sigma$ to
$a_k$ is given by
\mbox{$\varepsilon_{m_{\phi}}^k=\frac{1}{n}\frac{\phi'(a_k)}{\phi'(m_{\phi})}\frac{a_k}{m_{\phi}}$},
so that:

@@

The elasticity of the arithmetic mean $m$ to $a_k$ is
increasing in $a_k$ whereas the elasticity of $m_{\sigma}$ is
decreasing in $a_k$. A one-percent increase in the importations
from $k$ increases more the arithmetic mean of importations if
$k$ is initially an important importer, simply because $k$
represents a larger part of trade. But it is the opposite with
$m_{\sigma}$: the smaller the initial value of trade with $k$,
the bigger the impact of its growth on $m_{\sigma}$. Therefore,
Jacks and al.'s measure is biased toward the growth rates of
costs incurred with small trade partners.

>>>>

This is potentially an important result that sheds light on the globalization processes. However, that result is actually driven by an \textit{ad hoc}
method of aggregation that captures structure effects. The inferred difference between the two globalization comes from the fact that the dyads with the 
fastest growing trade in the first wave of globalization start with very small 
trade. This is not the case in the second wave of globalization.
Using a three-countries version of \cite{AW2003}'s model for
the aggregation of bilateral trade costs shows that there was no 
difference in nature between the two globalizations. More generally, we
cast doubt on the possibility to distinguish the effect of trade costs and economic expansion through an approach that relies only on the study of trade
flows. Deducing trade costs from trade flows, and then using
trade costs to explain trade flows is essentially a circular
reasoning. Therefore, Jacks et al.'s approach cannot be an
alternative to traditional investigations of impediments to
trade, such as commodity price gaps.

This paper starts by presenting Jacks et al.'s approach of the
measure of trade costs, and insists on its relevance at the
bilateral level to control for multilateral trade barriers in
gravity regressions (section \ref{1}). It then highlights that the
result on a difference of nature between the two globalizations
is paradoxical since it cannot be deduced from a comparison of the
evolution of openness ratios. Section \ref{2} shows that the
conclusion is only driven by the authors' \textit{ad hoc} aggregation
method. We propose a microfounded way to aggregate trade costs
and the puzzle fades away. Section \ref{3} explores the reasons
why Jacks et al.'s aggregation technique ends up providing
different results for the two globalizations. We argue that
what Jacks et al. misleadingly attribute to unequal trade costs
decreases between the two globalizations is instead a
difference in the distribution of trade growth over trading
partners.

@@

Jacks et al. first depart from \cite{AW2003} by eliminating the
multilateral resistance variable ($P_i P_j$) from the gravity equation.
Following \cite{NOVY}, bilateral trade barriers are not
considered relative to multilateral trade barriers modelized by
the price index, but to domestic trade costs. In this case,
trade flows are no longer compared to outputs, but to internal
trade $x_{ii}$\footnote{Due to data limitations, the authors
use the relation $x_{ii}=GDP_i-EXPORTS_i$ to get internal
trade. We follow them. Concerns about the fact that GDP is
measured in added-value and exportations as gross value are
addressed in appendix B of \cite{JMN2011}.}. The equation above
becomes:

@@

Our point is that a method that uses trade flows
alone can teach us much about their determinants. When
departing from \cite{AW2003}'s multilateral resistances, the
arbitrage condition Jacks et al. base their computation on is an
equality between a Marginal Rate of Substitution (MRS) and a
price ratio for a CES utility function, or more precisely the
product of two such equalities, one from the program of each
trade partner. Appendix \ref{A1} derives equation (\ref{keyeq})
this way without passing through multilateral resistance. The
inference of trade costs from trade flows is tantamount to
using the relation between quantities and prices given by a
demand curve derived from fully specified exogenous
preferences.

@@

Since the trade costs measure is assimilated to trade flows
relative to domestic activity, they simply restate
 an information already contained in any measure of
trade flows relative to economic size, like an openness ratio
defined for instance as the ratio of exportations to GDP.
However the main conclusion in Jacks et al.'s article, namely
that the first globalization was driven more by the decrease in
trade costs than the second one, requires the absence of circularity.
Indeed, when we decompose the level of
exportations of a country as the product of its GDP and its
openness ratio, we find results that are different from theirs (see table \ref{OR}).\footnote{We follow
the authors in interpreting log differences as percentages. However, one should keep in mind that given the
size of the changes, this is a very inexact approximation.} The increase in exportations is mainly explained by GDP growth for the large majority of the 27 countries in the
data. On average, this decomposition attributes 74\% (183/246) of
the growth in trade to the increase in GDP in the second
globalization and 60\% in the first globalization. This is very much in contrast to 
Jacks et al.\footnote{Contributions of growth in income
similarity and of change in multilateral factors are negative.} :

@@

Jacks et al.'s conclusion on a difference of nature between the
two globalizations is based on an aggregate trade costs
measure that provides a summary statistic of the
evolution of trade costs across all dyads in the sample, as
well as for different regions of the world. To move from
bilateral costs $\tau_{ij}$ to an aggregate measure of trade
costs, the authors use an arithmetic mean over dyads, weighted
by the sum of the GDP of the two trade partners.

@@

Let us note $\upsilon_{i}$ the corresponding trade cost faced
by country $i$ with all its trading partners in the data. Its
expression is given by Jacks et al.'s key equation
(\ref{keyeq}) applied to a three-countries model:

@@

Using this measure of trade costs, we provide the same
decomposition in four terms of the increase in trade flows
during the two globalizations as in the authors' article for
the countries with the most trade partners available in the
data (France (24), the UK (25), the USA (23)). We also provide
the unweighted and end-of-period-GDP-weighted averages for all
countries in the sample. Results are displayed in tables
\ref{glob1} and \ref{glob2} along with the results with the
Jacks et al.'s method.\footnote{They include unweighted and
end-of-period-GPD-weighted averages over country dyads of
bilateral trade costs, such as provided in \cite{JMN2011}. We
also display for France, the UK and the USA the trade cost
measure averaged over trading partners faced by one country, as
the results are presented (with a smaller data set) in
\cite{JMN2008, JMN2010}. We also provide both averages of this
measure over all countries in the data set.}\\

@@

Tables \ref{glob1} and \ref{glob2} highlight how much the decomposition
between the decrease in trade costs and the income growth depends on the aggregation technique. When we use the microfounded aggregation method,\footnote{We selected the (GDP-weighted) average of our measure in order
to allow a clear comparison with Jacks et al.'s results. There
is of course no rational for such a summary statistic, but the
results for France, the UK and the USA assure that the main
conclusion of this exercice does not depend on averaging over
countries.} 
growth in output is the main driving force behind both waves of growth in
international trade, contributing to about 56\% (203/365) in
the first wave and 66\% (361/545) in the second one. It is no coincidence
if these results are very similar to the ones in table \ref{OR}: 60\% (71/107) in the first globalization and 74\% (183/246) in the second one.
Insofar as a measure of
trade costs is defined residually as everything that explains
differences between domestic and international trade flows, it
is bound to reword the information contained in an openness
ratio.

@@

Our explanation of why Jacks et al.'s aggregate measure of
trade costs yields different conclusion 
from table \ref{OR} is because it is sensitive to structure
effects. To explain this idea, and for clarity purposes, let us consider
a word where partners are symmetric (domestic trade is equal in $i$ and $j$ and $x_ij$=$x_ji$). From equation \ref{keyeq}, we have:

@@

Trade costs imply that prices are specific to the country of
consumption. We note $p_k$ the supply price of the producer in
country $k$ net of trade costs, and $p_{ki}$ the price of
region $k$ goods for region $i$ consumers. We define
$t_{ki}=p_{ki}/p_{k}$ the trade costs factor between $k$ and
$i$. Jacks et al.'s variable of trade costs between $k$ and
$i$, $\tau_{ki}$ is then defined as the trade costs factor
between $k$ and $i$ relative to the domestic trade costs factor
$t_{ii}$: $\tau_{ki}=t_{ki}/t_{ii}-1$. In all that follows, we
use the same notations $\tau_{ki}$ for the geometric average of
$\tau_{ki}$ and $\tau_{ik}$. Symmetry of bilateral trade costs
does not need to be assumed.

@@

This is a static result. We are interested in the dynamic
behavior of trade flows. But Jacks et al.'s measure also
overweights initially small trade partners in terms of the
increase in trade flows. This can be seen by calculating the
elasticities to the importations from country $k$ ($a_k$) of
the arithmetic mean ($m$) and of Jacks et al.'s mean
($m_{\sigma}$) (keeping importations from other trade partners
$a_j$ constant). The elasticity of a generic mean $m_\sigma$ to
$a_k$ is given by
\mbox{$\varepsilon_{m_{\phi}}^k=\frac{1}{n}\frac{\phi'(a_k)}{\phi'(m_{\phi})}\frac{a_k}{m_{\phi}}$},
so that:

@@

The elasticity of the arithmetic mean $m$ to $a_k$ is
increasing in $a_k$ whereas the elasticity of $m_{\sigma}$ is
decreasing in $a_k$. A one-percent increase in the importations
from $k$ increases more the arithmetic mean of importations if
$k$ is initially an important importer, simply because $k$
represents a larger part of trade. But it is the opposite with
$m_{\sigma}$: the smaller the initial value of trade with $k$,
the bigger the impact of its growth on $m_{\sigma}$. Therefore,
Jacks et al.'s measure is biased toward the growth rates of
costs incurred with small trade partners.
